---
output: github_document
--- 
FortranwithR
======

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/prcwiek/FortranwithR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/prcwiek/FortranwithR/actions/workflows/R-CMD-check.yaml)
[![](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->

### Tutorial
Back to the future: How to use Fortran with R these days?

FortranwithR is an R package for learning how use Fortran with R.

### Installation

Get the development version from github:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("prcwiek/FortranwithR")
```

```{r}
library('FortranwithR')
```

### Contributors and wind measurement data example

Wind measurement data example was extracted from 
Wind resource data from the tall Risø met mast.
  
Hansen, Kurt Schaldemose; Vasiljevic, Nikola; Sørensen, Steen Arne (2021). 
Wind resource data from the tall Risø met mast. Technical University of Denmark. Dataset.  
https://doi.org/10.11583/DTU.14153204


