---
title: "Back to the future: How to use Fortran with R these days?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FortranwithR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: inline
---

```{r, label = "set options", include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE
)
```

```{r, echo=FALSE}
library(FortranwithR)
```


# Package FortranwithR

FortranwithR is an R package for learning how use Fortran with R.

During our walk through all steps will focus on several files with functon in R, one file in C, *fortranwithr_init.c*,  and one file in Fortran, *fortranwithr.f90* in the folder *src*.
The *Makevars* file was prepared to work with the package and it can be easily tailored to other
needs.

## Step 0 or the starting point
First, we will take a look at the template of our R package to get the first idea how the things work.  

The first function in our package, *hello_user.R* is as simple as possible. We don't pass any arguments and we don't process the output.

```{r 'First function', eval=FALSE}
hello_user <- function() {

  resp <- .C("hello_user_c")

}
```
It calls the function written in C in which we register the function name later used in the Fortran code.
Here it is *hello_user_c* with no arguments.  
  
extern void **hello_user_c()**  

{**"hello_user_c"**, (DL_FUNC) &hello_user_c, **0**}  

```{c, eval=FALSE}
#include <R.h>
#include <Rinternals.h>
#include <R_ext/Rdynload.h>

extern void hello_user_c();


static const R_CMethodDef cMethods[] = {
  {"hello_user_c", (DL_FUNC) &hello_user_c, 0},
  {NULL, NULL, 0}
};

void R_init_FortranwithR(DllInfo *dll)
{
  R_registerRoutines(dll, cMethods, NULL, NULL, NULL);
  R_useDynamicSymbols(dll, FALSE);
}

```

Finally, we write the Fortran part. We will place all functions in the *module*.
```{fortran, eval=FALSE}
module fortranwithr

  use, intrinsic  :: iso_c_binding

  implicit none
  private

contains

!**********************************************************
!* Hello UseR! 2026                                       *
!**********************************************************
  subroutine hello_user() bind(C, name = "hello_user_c")

      !print *, "Hello UseR! 2026"

  end subroutine hello_user

end module fortranwithr

```
Right now, the Fortran function doesn't do anything. If you uncomment the print statement, the package will still build, but you'll get warnings since writing to the console isn't allowed here.

## Step 1: Simple input and output
Let's see how can we pass arguments and get results using Fortran function. The example is trivial, but
it is just the beginning.  
First, we create a new package function in R with one input argument. However we will have to pass two
arguments to the .C function, to provide an input variable and receibe an output value.
```{r, my_sqrt, eval=FALSE}
#' My sqrt function
#'
#' @description
#' Step 1: Simple input and output
#' 
#' @param value numeric;  
#'
#' @return resp; a list with results
#'
#' @export
my_sqrt <- function(value) {

  resp <- .C("my_sqrt_c",
             value = as.numeric(value),
             result = as.numeric(value))
  
  resp$result
  
}

```
We have to register it in our C file
```{c, eval=FALSE}
#include <R.h>
#include <Rinternals.h>
#include <R_ext/Rdynload.h>

extern void hello_user_c();

extern void my_sqrt_c(double *value,
                      double *result);


static const R_CMethodDef cMethods[] = {
  {"hello_user_c", (DL_FUNC) &hello_user_c, 0},
  {"my_sqrt_c", (DL_FUNC) &my_sqrt_c, 2},
  {NULL, NULL, 0}
};

void R_init_FortranwithR(DllInfo *dll)
{
  R_registerRoutines(dll, cMethods, NULL, NULL, NULL);
  R_useDynamicSymbols(dll, FALSE);
}

```
And now we can write a function in Fortran. We will disgnuish between the input argument
we do not want to change **intent(in)** and the argument we want to change **intent(in out)**. We will use 
C variable types from the Fotran library *iso_c_binding* instead of standard Fortran variables.
```{fortran, eval=FALSE}
!**********************************************************
!* my_sqrt                                                *
!**********************************************************
  subroutine my_sqrt(value, result) bind(C, name = "my_sqrt_c")

      real(c_double), intent(in)        :: value
      real(c_double), intent(in out)    :: result

      result = sqrt(value)

  end subroutine my_sqrt
```
Let's see how it works.
```{r}
my_sqrt(4)
```
```{r}
my_sqrt(703)
```

## Wind measurement and wind shear
A short stop now as I would like to introduce you to the problem we want to solve. 

Let's talk about predicting wind conditions. Beforestarting a wind farms constructions,
investors want to knowthe future Annual Energy Production (AEP). The calculations are based on
costly wind resource measurements. We collect years of high-resolution data from measurement masts, typically recording 10-minute averages of wind speed, direction, pressure, and temperature.
It generates a massive volume of data.

However, we face a height gap. While measurement masts reach heights from 120 m to 160 m, today’s 
turbines often have hub heights around 180 m. The challenge lies in accurately predicting
wind behavior in that 'unmeasured' space above the mast.

### Wind profile
Wind speed changes with height and the wind profile is usually described by a power law:
$$u(z) = u(z_{ref}) \cdot\left(\frac{z}{z_{ref}}\right)^\alpha$$
where $u(z_{ref})$ is a wind speed at the reference height $z_{ref}$, $u(z)$ is a wind speed at
the height z and $\alpha$ is the power law exponent also called the shear coefficient.  

Example of three different wind profiles below.  
```{r wind_profiles, echo=FALSE, fig.show='hold', fig.width=6, fig.height=6}
v <- function(vref, z, zref, a ) {
  v <- vref * (z/zref)^a
}

df <- data.frame(h = seq(20, 200, 10))
df$v1<- v(5, df$h, 20, 0.10)
df$v2<- v(5, df$h, 20, 1/7)
df$v3<- v(5, df$h, 20, 0.20)

plot(df$v1, df$h, type = "l", col = "blue",
     xlim = c(5, 8), ylim = c(20, 200),
     xlab = "Wind speed (m/s)",
     ylab = "Height (m)")
lines(df$v2, df$h, type = "l", col = "green")
lines(df$v3, df$h, type = "l", col = "red")
legend("bottomright", 
       legend = c(round(0.1, 2), round(1/7, 2), round(0.2, 2)), 
       fill = c("blue", "green", "red"),
       title = "Alpha shear")

```
  
Usually we want to know wind profiles for 16 direction sectors, from 0 to 360 degrees. Additionally 
average wind speed and direction in each sector, number of valid data etc.

Typically, wind profiles are analyzed across 16 directional sectors to provide a full 360-degree characterization of the site. Key metrics for each sector include mean wind speeds, 
meand directions and data availability (the count of valid observations).

We will work with the data set from "Hansen, Kurt Schaldemose; Vasiljevic, Nikola; Sørensen, 
Steen Arne (2021). Wind resource data from the tall Risø met mast. Technical University of Denmark. 
Dataset. https://doi.org/10.11583/DTU.14153204.v1"

The dataset consists of 10 minutes average of:

- wind speeds at the heights of 44 m, 77 m, 125 m
- wind directions at the heights of 77 m, 125 m
- temperatures at the heiights of 3 m, 44 m and 118 m
- pressure at the height P8

```{r dataset}
dplyr::glimpse(winddata)
```

## Step 2: Short introduction to Fortran

### Fortran program structure

```{fortran example_structure, eval=FALSE}
program example_structure

    implicit none
  
    integer     :: i
  
    i = 1
    print *, i
  
end program example_structure
```

### Do loop
```{fortran do_loop_example, eval=FALSE}
program do_loop_example

    implicit none
  
    integer     :: i
  
    do i=1, 10, 2
        print *, i
    end do
  
end program do_loop_example
```

### Arrays
```{fortan arrays_example, eval=FALSE}
program arrays_example

    implicit none
  
    real        :: a(10)
    integer     :: i
  
    do i = 20, 11, -1
        a(i-10) = -i * 0.25
    end do

    do i= 1, 10
        print *, i, a(i)
    end do
  
end program arrays_example
```


### If
```{fortran if_example, eval=FALSE}
program if_example

    implicit none

    integer     :: i

    do i = 1, 10
        if (mod(i, 2) == 0) then
            print *, "Even number: ", i
        else 
            print *, "Odd number:  ", i
        end if
    end do
        
end program if_example
```

### Subroutines
```{fortran example_subroutine, eval=FALSE}
program subroutine_example

    implicit none

    integer     :: i

    do i = 1, 10
        call example(i)
    end do

contains

    subroutine example(n)

        integer, intent(in)     :: n

        if (mod(n, 2) == 0) then
            print *, "Even number: ", n
        else 
            print *, "Odd number:  ", n
        end if
        
    end subroutine example

end program subroutine_example
```

## Functions
```{fortran function_example, eval=FALSE}
program function_example

    implicit none

    real    :: a

    a =  my_sqrt(80.30)

    print *, a

contains

    real function my_sqrt(a) result(res)

        real, intent(in)    :: a

        res = sqrt(a)
    
    end function my_sqrt

end program function_example
```

### Modules
```{fortran module_example, eval=FALSE}
module my_module

    implicit none

contains

    pure real function my_sqrt(a) result(res)

        real, intent(in)    :: a

        res = sqrt(a)
    
    end function my_sqrt

    subroutine example(n)

        integer, intent(in)     :: n

        if (mod(n, 2) == 0) then
            print *, "Even number: ", n
        else 
            print *, "Odd number:  ", n
        end if
        
    end subroutine example

end module my_module

program module_example

    use my_module, only: my_sqrt
    
    implicit none

    real    :: a

    a =  my_sqrt(80.30)

    print *, a

end program module_example
```





## Step 3: Wind profile function in Fortran

## Step 4: Is it all worth it? Benchmarking

## Summary - where to go next


















